### 2014-10-30 Paper_v2.tex
Starting final version edits

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort&compress,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order modeling of blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
the human coagulation cascade.  
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
This formulation was more than an order of magnitude smaller than current coagulation models, 
because many of the mechanistic details of coagulation were encoded as logical rules.
We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
We then validated the model ensemble using thrombin data sets that were not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems.

\vspace{0.25in}
{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}

@@

%We chose coagulation because it is well studied, has a complex architecture, 
%and has an abundance of experimental data available for model identification and validation. 
%However, coagulation was just a proof-of concept to validate our approach; 
%

%Developing mathematical modeling tools to interrogate the molecular basis of human disease is one of the grand challenges facing systems biology.
%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
%Many have suggested that the integration of experimental and computational research is required to unravel
%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
%like thrombosis during surgery.
%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
%Development and validation of detailed network models requires a significant investment of time and material resources. 
%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.

%
%Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
%significant identification challenges.
%This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are common tools for modeling biochemical systems because of their ability to capture dynamics and encode mechanism.
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar \textit{et al.}, showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, it is not clear whether we actually need precise estimates for all model parameters. 
Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well-defined. 
Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships between model components using non-linear transfer functions, 
without requiring mechanism \citep{Morris:2010aa}. 
Despite their simplicity, logical models have captured rich behaviors in a variety of systems important to human health \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
However, modeling complex dynamics with logical models is challenging. 
Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
while missing or incomplete mechanistic knowledge can be approximated using a logical approach.

In this study, we developed a hybrid approach which combined ODE modeling with logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by TF/FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

%However, components could also correspond to non-modeled influences, categorial variables or other abstract quantities.
%In the current proof-of-principle model, 
%APC can inhibit amplification, and upstream of amplification thus APC activity will directly slow the rate of thrombin formation from the initiation trigger, and amplification itself.
%On the other hand, ATIII inhibits thrombin directly.

\subsection*{Formulation of reduced order coagulation models.}
We developed a reduced order extrinsic coagulation model to test our hybrid modeling approach (Fig. \ref{fig-network}).
A trigger event initiates thrombin formation (FIIa) from prothrombin (fII) through a lumped initiation step.
This step loosely represents the initial activation of thrombin by activated FXa.
Once activated, thrombin catalyzes its own formation (amplification step), and inhibition via the conversion of protein C to activated protein C (APC).
Antithrombin III (ATIII) inhibits amplification, while APC and tissue factor pathway inhibitor (TFPI) potentially inhibit both initiation and amplification. 
All initiation and inhibition processes, as well as the dependence of amplification upon other coagulation factors, was approximated using our rule-based approach (Fig. \ref{fig-control-schematic}). 
Individual regulatory contributions to the activity of pathway enzymes were integrated into control coefficients ($v$'s) using an integration rule (min/max).  
These control coefficients then modified the rates of model processes at each time step.
Hill-like transfer functions $0 \leq f\left(\mathcal{Z}\right) \leq 1$ quantified the contribution of components upon a target process. 
Components were either individual inhibitor or activator levels or some function of levels, e.g., the product of factor levels. 
In this study, $\mathcal{Z}$ corresponded to the abundance of individual inhibitors or activators, 
with the exception of the dependence of amplification upon specific coagulation factors (modeled as the product of factors). 
When a process was potentially sensitive to multiple inputs, logical integration rules were used to select which transfer functions influenced the process at any given time. 
In our proof of concept model, we used a winner takes all strategy; the maximum or minimum transfer function was selected at any given time step. 
However, other integration rules are certainly possible. 
Taken together, while the reduced order coagulation model encodes significant biological complexity, it is highly compact (consisting of only five differential equations). 
Thus, it will serve as an excellent proof of principle example to study the reduction of a highly complex human subsystem.

\subsection*{Identification of model parameters using particle swarm optimization.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
the residual. However, not all parameters were varied simultaneously. 
We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem, e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.

The reduced order coagulation model captured the role of initial prothrombin abundance, and the decay of the thrombin signal following from ATIII activity (Fig. \ref{fig-diff_fII_NO_APC}).
However, we systematically under-predicted the thrombin peak and the strength of ATIII inhibition in this training data set.
On the other hand, with fixed thrombin parameters, we captured peak thrombin values and the decay of the thrombin signal (at least for the 150\% fII case) in the presence
of both ATIII and the protein C pathway (Fig. \ref{fig-diff_fII_APC}). Lastly, we were unable to capture global differences in initiation time \textit{across} separate data 
sets with a single ensemble of model parameters. These differences likely resulted from normal experimental variability.
For example, different thrombin generation experiments within the 
training data (at the same physiological factor levels) had significantly different initiation times (data not shown). 
However, the inability to globally capture initiation time also highlighted a potential shortcoming of the initiation module within the model. 
To capture the variability in initiation time \textit{across} training data sets, we included a constant time-delay parameter ($T_{D}$) for each data group.
The delay parameter was constant within a data set, but allowed to vary \textit{across} training data sets. 
Introduction of the delay parameter allowed the model to simulate multiple training data sets using a single ensemble of model parameters.
Taken together, the model identification results suggested 
that our hybrid approach could reproduce a panel of thrombin generation data sets in the neighborhood of physiological factor and inhibitor concentrations.
However, it was unclear whether the reduced order model could predict new data, without updating the model parameters.

>>>>

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
\vspace{0.05in}
{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
Androgens, such as testosterone, are required for prostate cancer growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer. 
However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
CRPC is closely related to metastasis and decreased survival. 
In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
Additionally, data from four drug trials was also used to evaluate the model’s performance.
Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive. 
In a constant DHT environment sensitive species were conserved between the cell lines. 
These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for AIPC.

\vspace{0.5in}
{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}

@@

\section*{Introduction}

%
%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen. Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation. Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop. The model was validated with multiple experiments in LNCaP cell lines. In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.

Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen-dependent prostate cancer (ADPC) \cite{Huggins1967}. 
However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
Other approved treatments are non-specific to PCa. 
For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
advantage.

Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}. 
Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers. 
In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}). 
Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
In CRPC, AR signals in the absence of androgens. 
Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}. 
In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}. 
Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.

In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
We used this model to better understand which components and processes were differentially important in AD versus CR cells.
The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1). 
Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F. 
However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
Of these, 108 were larger than one standard deviation above the mean.
In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.

%(323 more and 287 less sensitive)
%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
%transition from androgen dependent to castration resistant prostate cancer.
%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
%that was an extension of our previous work .
%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
%First, we expanded the original model by adding ten  
%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
%ODE models typically contain a large number of unknown model parameters, which must be estimated  
%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
%All protein and RNA species were degraded with first order reaction rates. 
%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
%Relative concentration values were quantified 
%Training data came from experiments on LNCaP sublines. 
%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19

@@

\subsection*{Estimating an ensemble of prostate signaling models.} 
We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}). 
Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.5 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. ZZ). 
As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.

The ensemble of PCa models recapitulated training data in both AD and CR cell lines with only two experimentally justified parameter changes (Fig. \ref{fg:ColorPlot} and Fig. \ref{fg:Training}). 
Data from the LNCaP clones C-33 (dependent), C-51 (intermediate), and C-81 (resistant) \cite{Lin2000,Igawa2002,Horoszewicz1983} along with the CR LNCaP cell lines LNCaP-Rf \cite{Murillo2001}, LNCaP-AI \cite{Chen2011} and LNAI \cite{Graff2009} were used for model identification. 
To simulate the effective difference between LNCaP cell lines, the parameter controlling the maximum rate of PAcP gene expression was scaled by 0.1 and 0.5, respectively, for the C-81 and C-51 cell-lines compared to C-33. 
This modification was based upon steady-state PAcP data from the three LNCaP clones \cite{Lee2003}. 
Similarly, the expression of p16INK4 was adjusted in accordance with the study of Lu \textit{et al.} \citep{Lu1997}. 
These two parameters were the only adjustable parameter differences between AD and CR cells. 
To simulate an increased mTOR activation in the presence of a DHT stimulus, we added a first order activation term for mTOR activation with a DHT stimulus. Androgens have been shown to increase expression of proteins involved in cellular metabolism, which may lead to an increase in mTOR activation \cite{Xu2006}.
The model fit 36 of the 43 training objectives for greater than 40\% of the ensemble members (Fig. \ref{fg:ColorPlot}A). 
Conversely, only 10 of the 43 training objectives were captured with the random parameter control (Fig. \ref{fg:ColorPlot}B). 
The model captured the crosstalk between RTK activation and androgen action (Fig. \ref{fg:Training}). 
The model described DHT-induced PSA expression in both C-33 (Fig. \ref{fg:Training}A) and C-81 (Fig. \ref{fg:Training}B) cells. 
Interestingly, simulations with the HER2 inhibitor AG879 recapitulated decreased PSA expression in C-81 cells (Fig. \ref{fg:Training}C) in the absence of androgen stimulation.
AR action decreased the PAcP mRNA message (Fig. \ref{fg:Training}D), presumably leading to increased HER2 activity. 
The model also recapitulated the integration of androgen action with AR expression, G1/S cell cycle protein expression and AKT phosphorylation. 
For example, the model captured AR-induced AR expression following a DHT stimulus (Fig. \ref{fg:Training}H).
Conversely, the transcription factor E2F inhibits AR transcription in LNCaP cells (Fig. \ref{fg:Training}I). 
Other cell cycle proteins were also integrated with androgen action.
For example, the cyclin D1 abundance increased in CR compared to AD cells in the absence of androgen (Fig. \ref{fg:Training}E), while DHT induced p21Cip1 expression in C-33 cells (Fig. \ref{fg:Training}F).
The level of phosphorylated AKT was also increased in higher passage number cells (Fig. \ref{fg:Training}G). Taken together, [FINISH ME].
